MachLLensFinder - I: A Machine Learning Strong Lensing Identification Pipeline

### 2016-11-07 ms.tex
More text in results added

\author{Camille Avestruz\altaffilmark{1-3}\thanks{E-mail:
    avestruz@uchicago.edu}, Nan Li\altaffilmark{2,3,4}, and Matthew Lightman\altaffilmark{5} }

\affil{
$^1${Enrico Fermi Institute, The University of Chicago, Chicago, IL 60637 U.S.A.}\\
$^2${Kavli Institute for Cosmological Physics, The University of Chicago, Chicago, IL 60637 U.S.A.}\\
$^3${Department of Astronomy \& Astrophysics, The University of Chicago, Chicago, IL 60637 U.S.A.};
  \href{mailto:avestruz@uchicago.edu}{avestruz@uchicago.edu}\\
}

@@

%-------------------------------------------------%
\begin{abstract} 
  Gravitational lensing offers a direct probe of the underlying mass
  distribution of lensing systems, a window to the high redshift
  universe, and a geometric probe of cosmological models.  The advent
  of large scale surveys such as LSST and Euclid has prompted a need
  for automatic and efficient identification of strong lensing
  systems.  In this first paper of a series, we present LensFinderML,
  a strong lensing identification pipeline that will be publicly
  released as open source software.

Giant arcs trace... cosmology... mass distribution.  Advent of DES,
LSST, etc. need an efficient way to identify strong lens images.  We
present {\sc{MACHLLENSFINDER}} (MACHine Learning LENSFINDER), an open source
image processing and deep learning.  This first paper focuses on image
Processing Techniques for Giant Arc Identification Tests with mock
images from ... ...  Image processing improves by....
\end{abstract}
%-------------------------------------------------%

@@

%-------------------------------------------------%
\subsection{Image Processing Pipeline}
%-------------------------------------------------%

**Results of Grid search across the parameters of the pipeline**

%-------------------------------------------------%
\subsubsection{Histogram Oriented Gradients}
%-------------------------------------------------%
Histogram oriented gradients (HOG) is a feature extraction method that
computes centered horizontal and vertical gradients with no smoothing.
Originally created for human detection, .... This computes the
gradient orientation and magnitudes in an image by first dividing the
image into $n\times n$ blocks of X\% overlap.  The gradient
orientation is then quantized into m bins. ``(optional) global image
normalisation computing the gradient image in x and y computing
gradient histograms normalising across blocks flattening into a
feature vector''\citep{dalalandtriggs_05}.

@@

%-------------------------------------------------%
\subsection{Logistic Regression}
%-------------------------------------------------%

@@

%-------------------------------------------------%
\subsection{Deep Learning Algorithm}
%-------------------------------------------------%

%-------------------------------------------------%
\subsection{Mock Test Images}
%-------------------------------------------------%

%-------------------------------------------------%
\section{Results}
\label{sec:results}
%-------------------------------------------------%
%-------------------------------------------------%
\subsection{Optimized pipeline parameters}
%-------------------------------------------------%

\begin{table}
\caption{Grid Search of Pipeline Parameters}
\begin{center}
\begin{tabular}{cccccc}
\hline \\ [-0.2cm]
N$_{orient}$ & Pixels/Cell & Cells/Block & LogReg$_{\text{coeff}}$ & HOG size & Score \\ [0.2cm]
\hline \\ [-0.2cm]
\multicolumn{6}{c}{(a) HST-like data} \\ [0.2cm]
\hline \\ [-0.2cm]
9 & (4, 4) & (3, 3) & 50. & &$0.76181\pm0.00491$\\ [0.2cm]
9 & (4, 4) & (1, 1) & 50. & &$0.83403\pm0.00547$\\ [0.2cm]
9 & (8, 8) & (3, 3) & 50. & &$0.85417\pm0.00851$\\ [0.2cm] 
9 & (8, 8) & (1, 1) & 50. & &$0.86389\pm0.00687$\\ [0.2cm] 
9 & (16, 16) & (1, 1) & 50. & & $0.88542\pm0.00680$ \\ [0.2cm]
9 & (16, 16) & (3, 3) & 10. & &$0.89306\pm0.00687$ \\ [0.2cm]
8 & (16, 16) & (3, 3) & 50. & &$0.90000\pm0.00170$ \\ [0.2cm]
9 & (16, 16) & (3, 3) & 50. & &$0.90139\pm0.00804$ \\ [0.2cm]
\hline \\ [-0.2cm]
\multicolumn{6}{c}{(b) LSST-like data} \\ [0.2cm]
\hline \\ [-0.2cm]
& & & & &\\
& & & & &\\ [0.2cm]
\hline
\end{tabular}
\end{center}
\label{tab:gridsearch}
\tablecomments{Panel (a) shows the results of a grid search across a
  range of HOG parameters and regression coefficient with training and
  test sets of size 1440 with the HST-like mock dataset.  Panel (b)
  shows the corresponding results for training and test sets of size
  5760 with the LSST mock dataset.  We explore different HOG
  parameters in each dataset due to resolution and image size
  differences.}
\end{table}

@@

The number of orientations will determine the sampling of rounded
edges.  For example, if we only have two orientations, an arc-like
feature in a cell will appear as .... However, contributions from a
cluster or line-of-sight galaxy in the same cell will ... the
normalizations in the histogram.  The resolution of the overall image
will also limit the additional information that an increase in
$N_\text{orientations}$ will provide.  For the LSST-like data,
$N_\text{orientations}=9$ compared with $N_\text{orientations}=5$ only
results in a $???\%$ improvement.  In the case of the HST-like data,
$N_\text{orientations}=9$ compared with $N_\text{orientations}=5$
results in a $???\%$ improvement.

Note, we also include the size of the extracted features, e.g. the
length of the HOG.  Larger features will increase the amount of time
required to train a model, and also impact memory restrictions on the
size of the training set.  The logistic regression coefficient will
also increase the training time.  We discuss the score dependence on
logistic regression coefficient in Section\ref{sec:unregularizedfit}.

\subsection{Scores vs. regularization on larger train/test set with comparable grid search in parameterization}\label{sec:unregularizedfit}

\subsection{False positives/negatives and why... something about likely images given cosmology}

\subsection{ROC curve}

\subsection{Effects of BCG contamination}

>>>>

\author{Camille Avestruz\altaffilmark{1-3}\thanks{E-mail:
    avestruz@uchicago.edu}, Nan Li\altaffilmark{2-4}, and Matthew Lightman\altaffilmark{5} }

\affil{
$^1${Enrico Fermi Institute, The University of Chicago, Chicago, IL 60637 U.S.A.}\\
$^2${Kavli Institute for Cosmological Physics, The University of Chicago, Chicago, IL 60637 U.S.A.}\\
$^3${Department of Astronomy \& Astrophysics, The University of Chicago, Chicago, IL 60637 U.S.A.};\\
$^4${High Energy Physics Division, Argonne National Laboratory,
Lemont, IL 60439};\\
$^5${Digital Intelligence, JPMorgan Chase, Chicago, IL 60603 U.S.A.};\\
  \href{mailto:avestruz@uchicago.edu}{avestruz@uchicago.edu}\\
}

@@

%-------------------------------------------------%
\begin{abstract} 
  Gravitational lensing offers a direct probe of the underlying mass
  distribution of lensing systems, a window to the high redshift
  universe, and a geometric probe of cosmological models.  The advent
  of large scale surveys such as the Large Synoptic Sky Telescope and
  Euclid has prompted a need for automatic and efficient
  identification of strong lensing systems.  In this first paper of a
  series, we present LensFinderML, a strong lensing identification
  pipeline that will be publicly released as open source software.

Giant arcs trace... cosmology... mass distribution.  Advent of DES,
LSST, etc. need an efficient way to identify strong lens images.  We
present {\sc{MACHLLENSFINDER}} (MACHine Learning LENSFINDER), an open
source image processing and deep learning.  This first paper focuses
on image Processing Techniques for Giant Arc Identification Tests with
mock images from the Hubble Space Telescope and LSST.  Image
processing improves by....
\end{abstract}
%-------------------------------------------------%

@@

%-------------------------------------------------%
\subsection{Mock Images}
%-------------------------------------------------%

% Summary of HST and LSST images
We use mock {\em Hubble Space Telescope} (HST) and {\em Large Synoptic
  Sky Telescope} (LSST) images generated with PICS (Pipeline for
Images of Cosmological Strong lensing) \citep{li_etal16}.  These
images include cluster-member galaxies and foreground stars.  Half of
the images used are strong lensing systems.  Here, lensed galaxy
images are ray-traced images of actual galaxies extracted from deep
Hubble Space Telescope observations.  The final mock images are
``observed'' with a realistic point spread function corresponding to
modeled detector artifacts for bright stars.  See \cite{li_etal16} for
more details.

Figure ~\ref{fig:mockimages} illustrates a sample mock lensed and
unlensed images from each telescope.  From left to right, the top
panel images respectively correspond to a mock HST lensing and
non-lensing system. The mock HST images are $n_\text{pix}\times
n_\text{pix}=300\times300$ in size.  For the HST-like dataset, many
giant arcs are visually obvious.

The bottom panel images respectively correspond to a mock LSST lensing
and non-lensing system.  These images are $n_\text{pix}\times
n_\text{pix}=45\times45$.  The ground based noise and limited
resolution of the LSST-like dataset make visual giant arc
identification difficult.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]\label{fig:mockimages}
\begin{center}
\includegraphics[width=1\columnwidth]{figures/105_lensed_imgs_1_color.pdf}\hspace{-20pt}
\includegraphics[width=1\columnwidth]{figures/105_unlensed_imgs_1_color.pdf}\\
\includegraphics[width=1\columnwidth]{figures/lsst_9_lensed_imgs_1_color.pdf}\hspace{-20pt}
\includegraphics[width=1\columnwidth]{figures/lsst_9_unlensed_imgs_1_color.pdf}
\caption{Top panel: Example mock HST images, with $n_\text{pix}\times
  n_\text{pix}=300\times300$.  The left corresponds to a strong
  gravitational lensing system with a visibly obvious giant arc.  The
  right corresponds to a non-lensing system.  Bottom panel: Example
  mock LSST images, with $n_\text{pix}\times n_\text{pix}=45\times45$.
  The resolution and noise of a ground based telescope is noticeably
  worse.  The left corresponds to a strong gravitational lensing
  system, and the right to a non-lensing system.  However, visual
  identification of giant arcs in the LSST images is very difficult.}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-------------------------------------------------%
\subsection{Strong Arc Lensing Identification Pipeline}
%-------------------------------------------------%

To perform our analysis, we have used a variety of tools from {\em
  Scikit-learn} \citep{pedregosa_etal12}.  We outline the
identification pipeline in Figure~\ref{fig:pipeline}.  The first step
of our pipeline consists of a feature extraction stage, where our
feature vector is a {\em histogram of oriented gradients}
\citep{dalalandtriggs_05}.  We describe the method in
Section~\ref{sec:hog}.  We then use a machine learning algorithm to
train a classifier model on a subset of our images.  We have explored
both {\em Logistic Regression} (LR) and {\em Support Vector Machine}
(SVM) machine learning algorithms.  We test our trained model on an
independent subset of the images to assess the model performance.

Both the feature extrator, HOG, and the classifier contain a range of
parameters, which must be tested and optimized for peak model
performance.  We use {\em GridSearchCV} from {\em Scikit-learn} to
select appropriate parameters, and discuss these results in
Section~\ref{sec:gridsearch}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]\label{fig:pipeline}
\begin{center}
\includegraphics[scale=0.55]{figures/supervised_classification_workflow.png}
\caption{** Cartoon of pipeline ** Mention that median filtering generates
visual artifacts that confuse the model. http://www.nltk.org/book/ch06.html}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-------------------------------------------------%
\subsection{Feature Extractor: Histogram Oriented Gradients}\label{sec:hog}
%-------------------------------------------------%
Histogram oriented gradients (HOG) is a feature extraction method that
computes centered horizontal and vertical gradients with no smoothing.
Originally created for human detection, .... This computes the
gradient orientation and magnitudes in an image by first dividing the
image into $n\times n$ blocks of X\% overlap.  The gradient
orientation is then quantized into m bins. ``(optional) global image
normalisation computing the gradient image in x and y computing
gradient histograms normalising across blocks flattening into a
feature vector''\citep{dalalandtriggs_05}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]\label{fig:hogvisual}
\begin{center}
\includegraphics[width=1\columnwidth]{figures/105_lensed_imgs_1or5ppc32cpb1.pdf}\hspace{-20pt}
\includegraphics[width=1\columnwidth]{figures/105_lensed_imgs_1or9ppc16cpb1.pdf}\hspace{-20pt}
\caption{Visualized HOG feature vectors of the sample lensed HST-like
  image from Figure~\ref{fig:mockimages}.  Left: The corresponding HOG
  visualization with N$_\text{orient}=5$, (32, 32) pixels per cell,
  and (1, 1) cells per block.  Right: The HOG visualization of the
  same mock image with N$_\text{orient}=9$, (16, 16) pixels per cell,
  and (1, 1) cells per block.  Note, fewer pixels per cell will better
  resolve the arc feature.  The brightest lines align with prominent
  edges.}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-------------------------------------------------%
\subsection{Machine Learning Algorithm: Logistic Regression}
%-------------------------------------------------%

@@

%-------------------------------------------------%
\subsection{Machine Learning Algorithm: Support Vector Machine}
%-------------------------------------------------%

%-------------------------------------------------%
%\subsection{Deep Learning Algorithm}
%-------------------------------------------------%

%-------------------------------------------------%
\section{Results}
\label{sec:results}
%-------------------------------------------------%
%-------------------------------------------------%
\subsection{Optimized pipeline parameters}\label{sec:gridsearch}
%-------------------------------------------------%

\begin{table}
\caption{Grid Search of Pipeline Parameters}
\begin{center}
\begin{tabular}{cccccc}
\hline \\ [-0.2cm]
N$_{orient}$ & Pixels/Cell & Cells/Block & C$_{reg}$ & HOG size & Score \\ [0.2cm]
\hline \\ [-0.2cm]
\multicolumn{6}{c}{(a) HST-like data} \\ [0.2cm]
\hline \\ [-0.2cm]
9 & (4, 4) & (3, 3) & 50. & &$0.76181\pm0.00491$\\ [0.2cm]
9 & (4, 4) & (1, 1) & 50. & &$0.83403\pm0.00547$\\ [0.2cm]
9 & (8, 8) & (3, 3) & 50. & &$0.85417\pm0.00851$\\ [0.2cm] 
9 & (8, 8) & (1, 1) & 50. & &$0.86389\pm0.00687$\\ [0.2cm] 
9 & (16, 16) & (1, 1) & 50. & & $0.88542\pm0.00680$ \\ [0.2cm]
9 & (16, 16) & (3, 3) & 10. & &$0.89306\pm0.00687$ \\ [0.2cm]
8 & (16, 16) & (3, 3) & 50. & &$0.90000\pm0.00170$ \\ [0.2cm]
9 & (16, 16) & (3, 3) & 50. & &$0.90139\pm0.00804$ \\ [0.2cm]
\hline \\ [-0.2cm]
\multicolumn{6}{c}{(b) LSST-like data} \\ [0.2cm]
\hline \\ [-0.2cm]
& & & & &\\
& & & & &\\ [0.2cm]
\hline
\end{tabular}
\end{center}
\label{tab:gridsearch}
\tablecomments{Panel (a) shows the results of a grid search across a
  range of HOG parameters and regression coefficient with training and
  test sets of size 1440 with the HST-like mock dataset.  Panel (b)
  shows the corresponding results for training and test sets of size
  5760 with the LSST mock dataset.  We explore different HOG
  parameters in each dataset due to resolution and image size
  differences.}
\end{table}

@@

The number of orientations will determine the sampling of rounded
edges.  For example, if we only have two orientations, an arc-like
feature in a cell will appear as .... However, contributions from a
cluster or line-of-sight galaxy in the same cell will ... the
normalizations in the histogram.  The resolution of the overall image
will also limit the additional information that an increase in
$N_\text{orient}$ will provide.  For the LSST-like data,
$N_\text{orient}=9$ compared with $N_\text{orient}=5$ only
results in a $???\%$ improvement.  In the case of the HST-like data,
$N_\text{orient}=9$ compared with $N_\text{orient}=5$
results in a $???\%$ improvement.

Note, we also include the size of the extracted features, e.g. the
length of the HOG feature vector.  A longer feature vector will
increase the amount of time required to train the model.
Additionally, for fixed memory restrictions, there is a tradeoff
between the length of the feature vector and the size of the training
set.  The logistic regression coefficient will also increase the
training time, \todo{since larger coefficients correspond to a more
  singular matrix}.  We discuss the score dependence on logistic
regression coefficient in Section\ref{sec:unregularizedfit}.

%-------------------------------------------------%
\subsection{Effects of Regularization on Model Performance}\label{sec:unregularizedfit}
%-------------------------------------------------%

% Scores vs. regularization on larger train/test set with comparable
% grid search in parameterization

%-------------------------------------------------%
\subsection{Failed identifications}
%-------------------------------------------------%

%-------------------------------------------------%
\subsection{Receiver Operatic Characteristic}
%-------------------------------------------------%

In this section, we discuss the Receiver Operating Characteristic
(ROC) curve, which shows the true positive rate as a function of false
positive rate for a given model and test set.  The true positive rate
is defined as the number of true positives divided by the total number
of positives (completeness of the positive identifications).  The
false positive rate is defined as the number of false positives
divided by the total number of negatives (impurity of the
identifications).  The ROC curve illustrates the performance of our
trained model as we vary the discrimination threshold.

The classifier model assigns a score to each test image, which is a
probability that the image is a strong lensing system.  To construct
the ROC curve, the test images are ranked by probability and the
discrimination threshold varied.

For very high discrimination threshold, we have maximal purity, but a
low completeness (bottom left region of the ROC).  FOr a very low
discrimination threshold, we have maximal completeness, but minimal
purity (top right region of the ROC).  The ideal model would have a
data point on the ROC curve that sits on the top left corner,
i.e. $(x, y) = (0, 1)$.

In the context of strong lensing systems, we wish to maximize the true
positive rate so we have a representative count of the fraction of
strong lensing systems in an observed volume of the universe.  We also
want to minimize the false positive rate.  Positively identified
strong lensing systems will require expensive spectroscopic follow-up
for validation.  The steepness of the ROC curve indicates how well the
model will optimize the two, and we can characterize the steepness by
the area under the curve (AUC).  A perfect model would have an AUC of
1, with data points along $(x=0, 0\leq y\leq1)$, and $(0\leq x\leq 1,
y=1)$.

Figure~\ref{fig:ROC} shows the \todo{ROC curve for XYZ(???) models,
with the AUC value indicated in the legend.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]\label{fig:ROC}
\begin{center}
\includegraphics[width=\columnwidth]{figures/ROC.pdf}
\caption{ROC curve for a model trained on the HST-like data with the
  Logistic Regression algorithm.  Here, we use the parameters: CPB=(1,
  1), $N_\text{orien}=9$, PPC=(16, 16), $C_\text{reg}=100.$.  The AUC
  is 0.966.}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-------------------------------------------------%
\subsection{Effects of BCG contamination}
%-------------------------------------------------%

Previous works to identify strong lensing systems have often removed
the BCG from images to better identify lensed galaxies.  Here, we
